---
title: "Comparing Climate Data"
author: "Michelle Fink"
date: "November 26, 2018"
output:
  ioslides_presentation:
    fig_height: 5.5
    fig_retina: null
    fig_width: 8.5
    highlight: tango
    smaller: yes
    widescreen: yes
  beamer_presentation:
    highlight: tango
  slidy_presentation:
    highlight: tango
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## WorldClim vs. NEX-DCP30
<div class="columns-2">
### WorldClim (UC-Berkely)
- Monthly temperature and precipitation + 19 bioclimatic metrics
- 30 arc-second (~800m) resolution
- 'Current' interpolated from actual station data, averaged over 1960-1990 *more or less*
    + interpolated using ANUSPLIN algorithm
- 'Future' includes **2050** (2041-2060) and **2070** (2061-2080), for 17 GCMs & 4 RCPs
    + downscaled and 'calibrated' using the Current dataset
- Version 1 (2005), includes 'Current' and 'Future'
- Version 2 (2016) meant to correct issues with v.1 *But is for Current (1970-2000) only*

### NEX-DCP30 (NASA)
- Modeled (not from station data), monthly temperature and precipitation
- 30 arc-second (~800m) resolution
- Bias Corrected Spatially Downscaled method, 33 GCMs (or ensemble) & 4 RCPs
- 'Current' spans 1950 - 2005, all years individually available
- 'Future' spans 2006 - 2099, all years individually available
- From 2013
- Michelle can create any combination of metrics from any timeperiod or model(s)
</div>

## Visualizing Differences
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(plyr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(raster)

# Theme function -----------------------------------
gg.theme <- function(type=c("clean","noax")[1],useArial = F){
  # Changed 05/30/2018 [MMF] -
  # Original had code to force fonts on Mac
  # I don't think my tweaks quite work (on Windows) yet, but it doesn't break anything.
  require(ggplot2)
  if(useArial){
    bf_font=windowsFonts("sans")
  } else {bf_font=windowsFonts("serif")}

  switch(type,
         clean = theme_bw(base_size = 16, base_family=bf_font) +
           theme(axis.text.x     = element_text(size = 14),
                 axis.title.y    = element_text(vjust = +1.5),
                 panel.grid.major  = element_blank(),
                 panel.grid.minor  = element_blank(),
                 legend.background = element_blank(),
                 legend.key = element_blank(),
                 panel.border = element_blank(),
                 panel.background = element_blank(),
                 axis.line  = element_line(colour = "black")),

         noax = theme(line = element_blank(),
                      text  = element_blank(),
                      title = element_blank(),
                      plot.background = element_blank(),
                      panel.border = element_blank(),
                      panel.background = element_blank())
  )
}

#function to create geom_polygon calls
fill_viol<-function(gr.df,gr,qtile=NULL,probs){
  # SETUP VIOLIN QUANTILE PLOTS -----------------------------------
  # This is adapted from: http://stackoverflow.com/questions/22278951/combining-violin-plot-with-box-plot
  # Changed 05/30/2018 [MMF] -
  # - filter out NA values that are generated when populations are not the same size
  ifelse(is.null(qtile),{
    cuts <- cut(gr.df$y, breaks = quantile(gr.df$y, probs, na.rm=T, type=3, include.lowest = T, right = T), na.rm=T)},{
      cuts <- cut(gr.df$y, breaks = qtile, na.rm=T)
    }
  )
  quants <- mutate(gr.df,
                   x.l=x-violinwidth/2,
                   x.r=x+violinwidth/2,
                   cuts=cuts)

  quants <- filter(quants, !is.na(cuts))

  plotquants <- data.frame(x=c(quants$x.l,rev(quants$x.r)),
                           y=c(quants$y,rev(quants$y)),
                           id=c(quants$cuts,rev(quants$cuts)))

  #plotquants <- filter(plotquants, !is.na(id))

  #cut by quantile to create polygon id
  geom <- geom_polygon(aes(x=x,y=y,fill=factor(id)),data=plotquants,alpha=1,na.rm = T)

  return(list(quants=quants,plotquants=plotquants,geom=geom))
}

vioQtile <- function(gg=NULL,qtiles=NULL,probs=seq(0,1,.25),labels=paste(probs[-1]*100),withData=F,usecol=F){
  require(ggplot2)
  # SETUP VIOLIN QUANTILE PLOTS -----------------------------------
  # This is adapted from: http://stackoverflow.com/questions/22278951/combining-violin-plot-with-box-plot
  #
  # Changed:
  # - Deal with 'empty' quantile groups
  # - Deal with original data
  # - More input, more output
  # Changed 05/30/2018 [MMF] - minor aesthetic tweaks
  g.df <- ggplot_build(gg)$data[[1]]    # use ggbuild to get the outline co-ords

  ifelse(is.null(qtiles),{
    gg <- gg + lapply(unique(g.df$group), function(x) fill_viol(g.df[g.df$group==x, ],x,NULL,probs)$geom)},{
      gg <- gg + lapply(unique(g.df$group), function(x) fill_viol(g.df[g.df$group==x, ],x,qtiles[x, ],probs)$geom)}
  )

  ifelse(usecol,{
    gg <- gg +
      scale_fill_hue(name="Quantile",labels=labels,guide=guide_legend(reverse=T,label.position="right"),h=c(250,300)) +
      stat_summary(fun.y=mean, geom="point", size=5, color="grey50", shape=22, fill="red")},{
        gg <- gg +
          scale_fill_grey(name="Quantile",labels=labels,guide=guide_legend(reverse=T,label.position="right")) +
          stat_summary(fun.y=mean, geom="point", size=3.5, color="grey50", shape=22, fill="white")
      })

  if(withData){
    ifelse(is.null(qtiles),{
      ggData <- lapply(unique(g.df$group), function(x) fill_viol(g.df[g.df$group==x,],x,NULL,probs))},{
        ggData <- lapply(unique(g.df$group), function(x) fill_viol(g.df[g.df$group==x,],x,qtiles[x,],probs))
      }
    )
    return(list(ggGraph=gg,ggData=ggData))
  } else {
    return(gg)
  }
}

#------
# Bring in the 'historic' rasters
wc1_now <- raster("M:/GIS_Projects/BLM_ClimateChange/Source/current/current_bio05.tif")
wc2_now <- raster("D:/GIS/Climate/WORLDCLIM/wc2_wus_bio5.tif")
nex_now <- raster("M:/GIS_Projects/BLM_ClimateChange/Source/NEX_derived/historical_tasmax_NEXdcp30_season_3.tif")
nxh_now <- raster("D:/GIS/Projects/BLM_AdaptationStrat/Env_inputs/s3_tmax_MIROC_ESM.tif")

# Clip out Bent County for Current
wc1n <- getValuesBlock(wc1_now, row=338, nrows=74, col=691, ncols=78)
wc2n <- getValuesBlock(wc2_now, row=926, nrows=74, col=1382, ncols=78) #W. US
nexn <- getValuesBlock(nex_now, row=338, nrows=74, col=691, ncols=78)
nxhn <- getValuesBlock(nxh_now, row=368, nrows=74, col=1282, ncols=78) #PJ range

rasdf <- tibble("WC1_now" = wc1n, "WC2_now" = wc2n, "NEX_now" = nexn, "NXH_now" = nxhn)
xdf <- rasdf %>% gather(Model, MaxTemp, 1:4) #transpose columns to groups

# Bring in the 'future' rasters
wc1_future <- raster("M:/GIS_Projects/BLM_ClimateChange/Source/rcp85/rcp85_2050_bioclim_05.tif")
#wc2_future # there isn't one for WC2
nex_future <- raster("M:/GIS_Projects/BLM_ClimateChange/Source/NEX_derived/rcp85_tasmax_NEXdcp30_season_3.tif")
nxh_future <- raster("D:/GIS/Projects/BLM_AdaptationStrat/Env_inputs/Future_2050/s3_tmax_MIROC_ESM.tif")

# Clip out Bent County for Future
wc1f <- getValuesBlock(wc1_future, row=338, nrows=74, col=691, ncols=78)
wc2f <- seq(0.003465,20,0.003465)
nexf <- getValuesBlock(nex_future, row=338, nrows=74, col=691, ncols=78)
nxhf <- getValuesBlock(nxh_future, row=368, nrows=74, col=1282, ncols=78) #PJ range

rasdf2 <- tibble("WC1_2050" = wc1f, "WC2_2050" = wc2f, "NEX_2050" = nexf, "NXH_2050" = nxhf)
xdf2 <- rasdf2 %>% gather(Model, MaxTemp, 1:4) #transpose columns to groups
xdf$MaxTempF <- xdf2$MaxTemp # Add to the previous data.frame

LasAnimas <- c(32.17, 34.89, 33.61) #1971-2000 JJA mean max https://wrcc.dri.edu/cgi-bin/cliMAIN.pl?co4834

mytheme <- gg.theme("clean", useArial=T)
probs <- seq(0,1,.25)

g.pv <- ggplot(xdf,aes(x=Model,y=MaxTemp)) +
  geom_violin(aes(group=Model),scale="width",color="grey80",fill="grey80",trim=T, adjust=0.8)

g.pv0 <- vioQtile(g.pv)

g.pv0 + ggtitle("Summer maximum temperature (Bent Co)") + xlab("") + ylab("Celcius") +
  ylim(30,39) + labs(tag = "Las Animas 1971-2000 summer maximum") +
  geom_hline(aes(yintercept=mean(LasAnimas)), color="red") + mytheme +
  theme(legend.position=c(0.11,0.8), legend.text=element_text(size=16),
        legend.title=element_text(size=16), axis.text.x=element_text(size=18),
        axis.text.y=element_text(size=18), axis.title.y=element_text(size=18, vjust=2.6),
        title=element_text(size=20), plot.tag.position=c(0.5,0.465),
        plot.tag=element_text(size=16, color="red"))
```

## Visualizing Differences
```{r echo=FALSE, message=FALSE, warning=FALSE}
g.pv0 + ggtitle("Summer maximum temperature (Bent Co)") + xlab("") + ylab("Celcius") +
  ylim(30,39) +
  geom_violin(aes(y=MaxTempF, group=Model),scale="width",color="grey80",fill="purple",trim=T, adjust=0.8) +
  geom_hline(aes(yintercept=mean(LasAnimas)), color="red") + mytheme +
  theme(legend.position="none", axis.text.x=element_text(size=18),
        axis.text.y=element_text(size=18), axis.title.y=element_text(size=18, vjust=2.6),
        title=element_text(size=20))
```

## Visualizing Differences - A different county
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Clip out Morgan County for Current
wc1n <- getValuesBlock(wc1_now, row=67, nrows=63, col=601, ncols=81)
wc2n <- getValuesBlock(wc2_now, row=655, nrows=63, col=1292, ncols=81) #W. US
nexn <- getValuesBlock(nex_now, row=67, nrows=63, col=601, ncols=81)
nxhn <- getValuesBlock(nxh_now, row=97, nrows=63, col=1192, ncols=81) #PJ range

rasdf <- tibble("WC1_now" = wc1n, "WC2_now" = wc2n, "NEX_now" = nexn, "NXH_now" = nxhn)
xdf <- rasdf %>% gather(Model, MaxTemp, 1:4) #transpose columns to groups
Ft.Morgan <- c(28.61, 32.11, 31) #1971-2000 JJA mean max https://wrcc.dri.edu/cgi-bin/cliMAIN.pl?co3038

# Clip out Morgan County for Future
wc1f <- getValuesBlock(wc1_future, row=67, nrows=63, col=601, ncols=81)
wc2f <- seq(0,20,0.0039193)
nexf <- getValuesBlock(nex_future, row=67, nrows=63, col=601, ncols=81)
nxhf <- getValuesBlock(nxh_future, row=97, nrows=63, col=1192, ncols=81) #PJ range

rasdf2 <- tibble("WC1_2050" = wc1f, "WC2_2050" = wc2f, "NEX_2050" = nexf, "NXH_2050" = nxhf)
xdf2 <- rasdf2 %>% gather(Model, MaxTemp, 1:4) #transpose columns to groups
xdf$MaxTempF <- xdf2$MaxTemp # Add to the previous data.frame

g.pv <- ggplot(xdf,aes(x=Model,y=MaxTemp)) +
  geom_violin(aes(group=Model),scale="width",color="grey80",fill="grey80",trim=T, adjust=0.8)

g.pv0 <- vioQtile(g.pv)

g.pv0 + ggtitle("Summer maximum temperature (Morgan Co)") + xlab("") + ylab("Celcius") +
  ylim(29,37.5) + labs(tag = "Fort Morgan 1971-2000 summer maximum") +
  geom_hline(aes(yintercept=mean(Ft.Morgan)), color="red") + mytheme +
  theme(legend.position=c(0.11,0.8), legend.text=element_text(size=16),
        legend.title=element_text(size=16), axis.text.x=element_text(size=18),
        axis.text.y=element_text(size=18), axis.title.y=element_text(size=18, vjust=2.6),
        title=element_text(size=20), plot.tag.position=c(0.5,0.33),
        plot.tag=element_text(size=16, color="red"))
```

## Visualizing Differences - A different county
```{r echo=FALSE, message=FALSE, warning=FALSE}
g.pv0 + ggtitle("Summer maximum temperature (Morgan Co)") + xlab("") + ylab("Celcius") +
  ylim(29,37.5) +
  geom_violin(aes(y=MaxTempF, group=Model),scale="width",color="grey80",fill="purple",trim=T, adjust=0.8) +
  geom_hline(aes(yintercept=mean(Ft.Morgan)), color="red") + mytheme +
  theme(legend.position="none", axis.text.x=element_text(size=18),
        axis.text.y=element_text(size=18), axis.title.y=element_text(size=18, vjust=2.6),
        title=element_text(size=20))
```

## The Bioclimatic Variables
<div style="float: left; width: 50%;">
- BIO1 = Annual Mean Temperature
- BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))
- BIO3 = Isothermality (BIO2/BIO7) (* 100)
- BIO4 = Temperature Seasonality (standard deviation *100)
- BIO5 = Max Temperature of Warmest Month
- BIO6 = Min Temperature of Coldest Month
- BIO7 = Temperature Annual Range (BIO5-BIO6)
- BIO8 = Mean Temperature of Wettest Quarter
- BIO9 = Mean Temperature of Driest Quarter
- BIO10 = Mean Temperature of Warmest Quarter
- BIO11 = Mean Temperature of Coldest Quarter
</div>
<div style="float: right; width: 50%;">
- BIO12 = Annual Precipitation
- BIO13 = Precipitation of Wettest Month
- BIO14 = Precipitation of Driest Month
- BIO15 = Precipitation Seasonality (Coefficient of Variation)
- BIO16 = Precipitation of Wettest Quarter
- BIO17 = Precipitation of Driest Quarter
- BIO18 = Precipitation of Warmest Quarter
- BIO19 = Precipitation of Coldest Quarter
</div>